*Supplementary Material 2*

### 2022-04-01 ms/si.Rmd
Update MS

# Specify path to FTOL cache
# FIXME: change to ftol_cache when ready
ftol_cache <- here::here("ftol_cache")

@@

# Make table of Pteridocat genera not in PPGI
pteridocat_not_in_ppgi <-
  accepted_species %>%
  count(genus, name = "n_species") %>%
  left_join(select(ppgi_taxonomy_mod, genus, notes), by = "genus") %>%
  assert(not_na, notes) %>%
  assert(is_uniq, genus) %>%
  filter(notes != "original PPGI 2016") %>%
  transmute(
    diff = "Pteridocat not in PPGI",
    genus,
    n_species
  ) %>%
  # Add notes
  left_join(
    select(pteridocat_ppgi_diff_notes, -n_species),
    by = c("diff", "genus")
  ) %>%
  assert(not_na, note)

# Make table of PPGI genera not in Pteridocat
ppgi_not_in_pteridocat <-
  ppgi_taxonomy_mod %>%
  filter(notes == "original PPGI 2016",
         class == "Polypodiopsida") %>%
  anti_join(accepted_species, by = "genus") %>%
  count(genus) %>%
  transmute(
    diff = "PPGI not in Pteridocat",
    genus
  ) %>%
  # Add notes
  left_join(
    pteridocat_ppgi_diff_notes, by = c("diff", "genus")
  ) %>%
  assert(not_na, note)

@@

**`r s_table("genera_diff")`**.
Differences in genera included as accepted in Pteridocat taxonomic database (https://github.com/pteridophy/pteridocat) vs. PPGI [@PteridophytePhylogenyGroupI2016] (`r result_file(s_table("genera_diff"), "csv") %>% fs::path_file()`).
"diff" is type of difference (Pteridocat genus not in PPGI or PPGI genus not in Pteridocat).
"n_species" is the number of accepted species in each genus (Pteridocat) or estimated number of species (PPGI).
References in "notes" column provided at end of Supplementary Materials.
"World Ferns" refers to @Hassler2022, which was modified to create Pteridocat.

@@

**`r s_table("fossils")`**.
Fossils used as calibration points for molecular dating (`r result_file(s_table("fossils"), "csv") %>% fs::path_file()`).
"n_fos" is the fossil ID number used in our fossil database (https://github.com/pteridophy/fossils).
"minimum_age" (Ma) is the age used for calibration; "lower" is the lower limit of the oldest stratigraphic age for the fossil; "upper" is the upper limit of the oldest stratigraphic age for the fossil.
For detailed notes and full references for each fossil, see Appendix S2.

@@

**`r s_table("accs")`**.
GenBank accessions used in this study (`r result_file(s_table("accs"), "csv") %>% fs::path_file()`).
"species" is the species name as it appears in the phylogenetic tree.
"seq_len" is the total number of bases excluding missing bases ("?", "N", or "-") and only includes bases belonging to loci used in this study (not the entire accession).
"sci_name" is the scientific name in the Pteridocat taxonomic database.
"ncbi_name" is the scientific name used in the NCBI taxonomic database.
"ncbi_taxid" is the NCBI taxonomy database unique identifier.
"outgroup_status" is "in" for ingroup taxa (ferns) and "out" for outgroup taxa.

@@

**`r s_figure("fossil_map")`**.
Maximum likelihood fern phylongey with node labels indicating placement of fossil constraints (`r result_file(s_figure("fossil_map"), "pdf") %>% fs::path_file()`).
Number for each constraint is age in Ma.

@@

# Make tibble for printing BS values
bs_tibble <- make_bs_tbl(phy)

# Make plot
# optional: round version
bs_plot_round <- 
  ggtree(
    phy,
    layout = "circular",
    branch.length = "none", 
    size = 0.1, color = "grey30") %<+% bs_tibble +
  # Add node points
  geom_nodepoint(
    aes(color = bs), size = 0.2) +
  scale_color_viridis_c(
    "BS",
    option = "B", na.value = "transparent",
    breaks = 0.25*0:4, labels = percent(0.25*0:4))

# Make plot
bs_plot <- 
  ggtree(
    phy,
    size = 0.1, color = "grey30") %<+% bs_tibble +
  # Add node points
  geom_nodepoint(
    aes(color = bs), size = 0.1) +
  scale_color_viridis_c(
    "BS",
    option = "B", na.value = "transparent",
    breaks = 0.25*0:4, labels = percent(0.25*0:4)) +
  # Add tip labels
  geom_tiplab(aes(label = label), size = 0.5, color = "grey30") +
  theme(legend.position = "bottom")
```

```{r bs-save, eval = params$doc_type == "doc"}
ggsave(
  plot = bs_plot,
  file = (result_file(s_figure("bs"), "pdf")),
  height = 60, width = 8, units = "in",
  limitsize = FALSE)
```

**`r s_figure("bs")`**.
Consensus phylogeny (Sanger dataset) with node labels showing bootstrap support (%) (`r result_file(s_figure("bs"), "pdf") %>% fs::path_file()`).
The backbone (plastome) phylogeny was used as a constraint tree, so all clades present in the backbone phylogeny tree receive 100% bootstrap support.

@@

# Format names of papers for printing
du_ages_sampling <-
  tribble(
    ~source, ~year, ~label,
    "schuettpelz", 2009, "Schuettpelz & Pryer 2009",
    "rothfels", 2015, "Rothfels et al. 2015",
    "ts", 2016, "Testo and Sundue 2016",
    "current", 2021, "This study",
  ) %>%
  bind_rows(
    du_schemes %>%
      mutate(
        year = 2020 + 0.1*seq_along(du_schemes$scheme),
        label = paste("Du et al. 2021 scheme", code)
      ) %>%
      transmute(source = scheme, year, label)
  ) %>%
  mutate(source = fct_reorder(source, year))

@@

**`r s_figure("du-ages")`**.
Stem age of fern families (Ma) estimated by selected studies, only including families that were sampled across all studies.
For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a; @Du2021].
For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].
@Du2021 used six different dating schemes; for details, see that paper.
Codes in parentheses after family names indicate major clade as in `r figure("ftol")`.
Period name abbreviations as follows: FIXME.

@@

```{r comp-fossils, fig.width = 7, fig.height = 9, eval = params$doc_type != "manual"}
ts_fossil_calibration_tips %>%
  select(minimum_age, node_calibrated) %>%
  mutate(dataset = "ts") %>%
  bind_rows(
    con_fossil_calibration_tips %>%
      select(minimum_age, node_calibrated) %>%
      mutate(dataset = "current")
  ) %>%
  mutate(node_calibrated = fct_reorder(node_calibrated, minimum_age)) %>%
  ggplot(aes(x = minimum_age, y = node_calibrated, fill = dataset)) +
  geom_col(position = position_dodge()) +
  scale_x_continuous("Age (millions of years)", expand = expansion(mult = c(0,0.1))) +
  theme_bw() +
  scale_fill_manual(
    values = c(current = "grey30", ts = "grey80"),
    labels = c(current = "This Study", ts = "Testo and Sundue 2016")
  ) +
  theme(
    axis.title.y = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

@@

# Make plot
ggplot(all_family_stem_ages) +
  geom_pointrange(
    aes(
      x = median, xmin = low, xmax = high,
      y = family,
      color = author, shape = author),
    position = position_dodge(width = 0.5)) +
  scale_shape_manual(
    values = ages_shapes,
    limits = ages_sampling$source, 
    labels = ages_sampling$label) +
  scale_color_manual(
    values = ages_colors,
    limits = ages_sampling$source, 
    labels = ages_sampling$label) +
  scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
  scale_x_reverse("Age (millions of years)", expand = c(0,0)) +
  coord_geo(
    dat = list("periods", "eras"),
    pos = list("b", "b"),
    abbrv = list(TRUE, FALSE),
    xlim = c(500, 0),
    fill = "transparent",
    color = "grey20",
    lab_color = "grey30",
    height = unit(1, "lines"),
    expand = TRUE,
    size = 3) +
  guides(shape = guide_legend(nrow = 2)) +
  theme_bw() +
  theme(
    axis.title.y = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

@@

**`r s_figure("ltt")`**.
Lineage through time (LTT) plots for the fern order Polypodiales (black; this study) and angiosperms [colored lines\; @Li2019; @Ramirez-Barahona2020].
Angiosperm datasets labeled with "CC", "RC", or "UC" correspond to the six analysis schemes of @Ramirez-Barahona2020.
Log number of lineages shown as percent of total to enable comparison across trees.

>>>>

# Specify path to FTOL cache
ftol_cache <- here::here("ftol_cache")

@@

# Make table of pteridocat genera not in PPGI
pteridocat_not_in_ppgi <-
  accepted_species %>%
  count(genus, name = "n_species") %>%
  left_join(select(ppgi_taxonomy_mod, genus, notes), by = "genus") %>%
  assert(not_na, notes) %>%
  assert(is_uniq, genus) %>%
  filter(notes != "original PPGI 2016") %>%
  transmute(
    diff = "pteridocat not in PPGI",
    genus,
    n_species
  ) %>%
  # Add notes
  left_join(
    select(pteridocat_ppgi_diff_notes, -n_species),
    by = c("diff", "genus")
  ) %>%
  assert(not_na, note)

# Make table of PPGI genera not in pteridocat
ppgi_not_in_pteridocat <-
  ppgi_taxonomy_mod %>%
  filter(notes == "original PPGI 2016",
         class == "Polypodiopsida") %>%
  anti_join(accepted_species, by = "genus") %>%
  count(genus) %>%
  transmute(
    diff = "PPGI not in pteridocat",
    genus
  ) %>%
  # Add notes
  left_join(
    pteridocat_ppgi_diff_notes, by = c("diff", "genus")
  ) %>%
  assert(not_na, note)

@@

**`r s_table("genera_diff")`**.
Differences in genera included as accepted in "pteridocat" taxonomic database v0.0.1 (https://github.com/fernphy/pteridocat) vs. PPGI [@PteridophytePhylogenyGroupI2016] (`r result_file(s_table("genera_diff"), "csv") %>% fs::path_file()`).
"diff" is type of difference (pteridocat genus not in PPGI or PPGI genus not in pteridocat).
"n_species" is the number of accepted species in each genus (pteridocat) or estimated number of species (PPGI).
References in "notes" column provided at end of Supplementary Materials.
"World Ferns" refers to @Hassler2022, which was modified to create pteridocat.

@@

**`r s_table("fossils")`**.
Fossils used as calibration points for molecular dating (`r result_file(s_table("fossils"), "csv") %>% fs::path_file()`).
"n_fos" is the fossil ID number used in the "ferncal" database v1.0.0 (https://github.com/fernphy/ferncal).
"minimum_age" is the age used for calibration (Ma); "lower" is the lower limit of the oldest stratigraphic age for the fossil (Ma); "upper" is the upper limit of the oldest stratigraphic age for the fossil (Ma).
For detailed notes and full references for each fossil, see Appendix S2.

@@

**`r s_table("accs")`**.
GenBank accessions used in this study (`r result_file(s_table("accs"), "csv") %>% fs::path_file()`).
"species" is the species name as it appears in the phylogenetic tree.
"seq_len" is the total number of bases excluding missing bases ("?", "N", or "-") and only includes bases belonging to loci used in this study (not the entire accession).
"sci_name" is the scientific name in the pteridocat taxonomic database.
"ncbi_name" is the scientific name used in the NCBI taxonomic database version 2022-02-01.
"ncbi_taxid" is the NCBI taxonomy database unique identifier.
"outgroup_status" is "in" for ingroup taxa (ferns) and "out" for outgroup taxa.

@@

**`r s_figure("fossil_map")`**.
Consensus phylogeny (Sanger dataset) with node labels indicating placement of fossil constraints (`r result_file(s_figure("fossil_map"), "pdf") %>% fs::path_file()`).
Number for each constraint is age in Ma.

@@

# Make tibble for printing BS values
bs_tibble <- make_bs_tbl(phy) %>%
  mutate(bs = bs*0.01) # so percentages print correctly

# Make plot
bs_plot <- 
  ggtree(
    phy,
    size = 0.1, color = "grey30") %<+% bs_tibble +
  # Add node points
  geom_nodepoint(
    aes(color = bs), size = 0.1) +
  scale_color_viridis_c(
    "BS",
    option = "B", na.value = "transparent",
    labels = scales::percent
    ) +
  # Add tip labels
  geom_tiplab(aes(label = label), size = 0.5, color = "grey30") +
  theme(legend.position = "bottom")
```

```{r bs-save, eval = params$doc_type == "doc"}
ggsave(
  plot = bs_plot,
  file = result_file(s_figure("bs"), "pdf"),
  height = 60, width = 8, units = "in",
  limitsize = FALSE)
```

**`r s_figure("bs")`**.
Consensus phylogeny (Sanger dataset) with node labels showing bootstrap support (%) (`r result_file(s_figure("bs"), "pdf") %>% fs::path_file()`).
Darker colors indicate lower support.
The backbone (plastome) phylogeny was used as a constraint tree, so all clades present in the backbone phylogeny tree receive 100% bootstrap support.

@@

# Format names of papers for printing
du_ages_sampling <-
  tribble(
    ~source, ~year, ~label,
    "schuettpelz", 2009, "Schuettpelz and Pryer 2009",
    "rothfels", 2015, "Rothfels et al. 2015",
    "ts", 2016, "Testo and Sundue 2016",
    "current", 2021, "This study",
  ) %>%
  bind_rows(
    du_schemes %>%
      mutate(
        year = 2020 + 0.1*seq_along(du_schemes$scheme),
        label = paste("Du et al. 2021 scheme", code)
      ) %>%
      transmute(source = scheme, year, label)
  ) %>%
  mutate(source = fct_reorder(source, year))

@@

**`r s_figure("du-ages")`**.
Stem age of fern families (Ma) estimated by selected studies, only including families that were sampled across all studies.
For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a; @Du2021].
For other studies, point indicates best (most likely) estimate [@Schuettpelz2009\; this study].
@Du2021 used six different dating schemes; for details, see that paper.
Codes in parentheses after family names indicate major clade as in `r figure("ftol")`.
Period name abbreviations as in `r figure("divtimes")`.

@@

```{r comp-fossils, fig.width = 7, fig.height = 9, eval = params$doc_type != "manual"}
ts_fossil_calibration_tips %>%
  select(minimum_age, node_calibrated) %>%
  mutate(dataset = "ts") %>%
  bind_rows(
    con_fossil_calibration_tips %>%
      select(minimum_age, node_calibrated) %>%
      mutate(dataset = "current")
  ) %>%
  mutate(node_calibrated = fct_reorder(node_calibrated, minimum_age)) %>%
  ggplot(aes(x = minimum_age, y = node_calibrated, fill = dataset)) +
  geom_col(position = position_dodge()) +
  scale_x_continuous("Age (millions of years)", expand = expansion(mult = c(0,0.1))) +
  theme_bw() +
  scale_fill_manual(
    values = c(current = "grey30", ts = "grey80"),
    labels = c(current = "This Study", ts = "Testo and Sundue 2016")
  ) +
  theme(
    axis.title.y = element_blank(),
    legend.title = element_blank(),
    axis.text.y = element_text(size = 6),
    legend.position = "bottom"
  )
```

@@

# Make plot
ggplot(all_family_stem_ages) +
  geom_pointrange(
    aes(
      x = median, xmin = low, xmax = high,
      y = family,
      color = author, shape = author),
    position = position_dodge(width = 0.5)) +
  scale_shape_manual(
    values = ages_shapes[c("ts", "current")],
    limits = c("ts", "current"), 
    labels = ages_sampling %>% 
      filter(source %in% c("current", "ts")) %>%
      arrange(source) %>%
      pull(label)) +
  scale_color_manual(
    values = ages_colors[c("ts", "current")],
    limits = c("ts", "current"), 
    labels = ages_sampling %>% 
      filter(source %in% c("current", "ts")) %>%
      arrange(source) %>%
      pull(label)) +
  scale_y_discrete(labels = family_labs, expand = expansion(add = 0.7)) +
  scale_x_reverse("Age (millions of years)", expand = c(0,0)) +
  coord_geo(
    dat = list("periods", "eras"),
    pos = list("b", "b"),
    abbrv = list(TRUE, FALSE),
    xlim = c(500, 0),
    fill = "transparent",
    color = "grey20",
    lab_color = "grey30",
    height = unit(1, "lines"),
    expand = TRUE,
    size = 3) +
  guides(shape = guide_legend(nrow = 2)) +
  theme_bw() +
  theme(
    axis.title.y = element_blank(),
    legend.title = element_blank(),
    legend.position = "bottom"
  )
```

@@

**`r s_figure("ltt")`**.
Lineage through time (LTT) plots for the fern order Polypodiales (black; this study) and angiosperms [colored lines\; @Li2019; @Ramirez-Barahona2020].
Angiosperm datasets labeled with "CC", "RC", or "UC" correspond to the six analysis schemes of @Ramirez-Barahona2020; for details, see that paper.
Log number of lineages shown as percent of total to enable comparison across trees.

### 2022-04-01 ms/si2.Rmd
Update MS

These data are exported from the FernCal v1.0 database (https://github.com/fernphy/ferncal), which includes `r nrow(fossil_ferns_raw)` records of fossils to use as calibration points for molecular dating of ferns.

@@

Taxonomy of extant taxa follows pteridocat v1.0 database (https://github.com/pteridophy/pteridocat).

>>>>

These data are exported from the FernCal v1.0.0 database (https://github.com/fernphy/ferncal), which includes records of fossils to use as calibration points for molecular dating of ferns.

@@

Taxonomy of extant taxa follows pteridocat v0.0.1 database (https://github.com/fernphy/pteridocat).
